---
title: "Data Management Assignment"
author: 
date: "September 23, 2015"
output: html_document
---

```{r, echo = FALSE, message = FALSE}
knitr::opts_chunk$set(comment = NA, message = FALSE, warning = FALSE, fig.align = "center")
```

### Purpose of Study: 

The present study will examine young adults from the National Epidemiologic Survey of Alcohol and Related Conditions (NESARC). The goals of the analysis will include 1) establishing the relationship between alcohol related problems and mental health issues(Depression, anxiety, and panic disorder) 2) determining whether or not the relationship between mental health issues and alcohol related problems is effected by gender.

### Variables:

Variables from NESARC that will be used include: `MAJORDEPLIFE` (has subject experienced a major depression?), `GENAXLIFE` (has subject expereinced an anxiety disorder in their lifetime), `PANLIFE` (has subject expereinced a panic disorder without agoraphobia?), `SEX` (is the subject male or female?), and `ALCABDEPP12DX` (has the subject abused, depended on, or not consumed any alcohol prior to the last 12 months).

### Data Management

First, the data is placed on the search path using the `PDS` package. The variables of interest are selected and stored in the data frame `nesarc.sub` using the `select` function from the `dplyr` package.

```{r}
library(PDS)
library(dplyr)
nesarc.sub <- NESARC %>% 
  select(MAJORDEPLIFE, GENAXLIFE, PANLIFE, SEX, ALCABDEPP12DX)
dim(nesarc.sub)
summary(nesarc.sub)
```

Running `summary` on the `nesarc.sub` reveals some non-obvious categories for the factors `CHECK321` and `S3AQ3B1`.  Reviewing the Code Book (HINT: Inside a pdf use `shift F` (Windows) or `command F` (Mac) then type the variable name inside the box to find the variable in the pdf), it is noted that a 9 is used to indicate missing values for `S3AQ3B1` and a 99 is used to indicate missing values for `S3AQ3C1`.

### Coding missing values

For my project the variables that I have chosen do not contain any missing variables. They are yes/no, or male/female. The alcohol dependence/abuse variable has four responses which are no alcohol diagnoses, alcohol abuse only, alcohol dependence only, and then alcohol abuse and dependence both. Therefore, none of my criteria has missing code values. 

### Labeling Variables

In the next section of code, variables are labeled and levels of factors are given informative labels.  The order of the levels is also rearranged for the variables `S3AQ3B1`, `TAB12MDX`, and `SEX`. 

```{r}
library(Hmisc)
label(nesarc.sub$MAJORDEPLIFE) <- "Major Depression in Lifetime"
label(nesarc.sub$GENAXLIFE) <- "Generealized Anxiety Disorder in Lifetime"
label(nesarc.sub$PANLIFE) <- "Panic Disorder Without Agoraphobia"
label(nesarc.sub$SEX) <- "Gender"
label(nesarc.sub$ALCABDEPP12DX) <- "Alcohol Abuse/Dependence Prior To the last 12 months"
nesarc.sub$MAJORDEPLIFE <- factor(nesarc.sub$MAJORDEPLIFE, 
                              labels = c("No Depression", "Yes Depression"))
nesarc.sub$GENAXLIFE <- factor(nesarc.sub$GENAXLIFE, 
                         labels = c("No Anxiety", "Yes Anxiety"))
nesarc.sub$PANLIFE <- factor(nesarc.sub$PANLIFE, 
                         levels = c("No Panic Disorder", "Yes Panic Disorder"))
nesarc.sub$SEX <- factor(nesarc.sub$SEX, labels = c("Male", "Female"))
table(nesarc.sub$SEX)
nesarc.sub$SEX <- factor(nesarc.sub$SEX, levels = c("Female", "Male"))
nesarc.sub$ALCABDEPP12DX <- factor(nesarc.sub$ALCABDEPP12DX, 
                         labels = c("No alcohol diagnoses", "Alcohol abuse only", "Alcohol dependence only", "Alcohol abuse and dependence"))
summary(nesarc.sub)
```

### Renaming Variables

```{r}
nesarc.sub <- nesarc.sub %>% 
  rename(Depression=MAJORDEPLIFE, Anxiety=GENAXLIFE, PanicDisorder=PANLIFE, Gender=SEX, AlcoholDependencyAndAbuse=ALCABDEPP12DX)
head(nesarc.sub)
```

### Creating Tables

I created frequency tables for the variables `AlcoholDependencyAndAbuse`,  `Depression`, and `Anxiety`.

```{r}
T1 <- xtabs(~AlcoholDependencyAndAbuse, data = nesarc.sub)
T2 <- xtabs(~Depression, data = nesarc.sub)
T3 <- xtabs(~Anxiety, data = nesarc.sub)
table(nesarc.sub$AlcoholDependencyandAbuse)
T1
T2
T3
```

In the data frame `nesarc.sub`, there are `r T1[1]` nicotine dependent subjects and `r T1[2]` subjects that are not nicotine dependent.  A small number of smokers (`r T2[4]`) smoke over 600 cigarettes per month. Most of the subjects in `nesarc.sub` are daily smokers (`r T3[6]`) with the remainder distributed uniformly across the first five levels of `SmokingFreq`.

### Graphing Frequency Tables

The barplots are all created with the package `ggplot2`.  The barplots start with the defaults for the `geom_bar` and add more detail to the plot with each graph.

```{r}
library(ggplot2)
g1 <- ggplot(data = nesarc.sub, aes(x = AlcoholDependencyAndAbuse)) + 
  geom_bar() +
  labs(x="Dependency/Abuse Of Alcohol", y="Total Number", title="Dependency Of Alcohol of Participants")
g1
```

```{r}
g2 <- ggplot(data = nesarc.sub, aes(x = Depression)) +
  labs(x= "Diagnosed with Depression(Lifetime", y="Total Number", Title= "Subjects Diagnosed With Depression") +
  geom_bar(fill = "SpringGreen2")
g2
```

```{r}
g3 <- ggplot(data = nesarc.sub, aes(Anxiety)) +
  labs(x = "Diagnosed With Anxiety", y = "Total Number", title = "Participants Diagnosed With Anxiety ") +
  geom_bar(fill = "maroon2") 
g3
```

